Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-10-03 main.tex
Fix probability example
Addressed Leila's comments.

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage[natbib=true,style=numeric,sorting=none]{biblatex}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\addbibresource{references.bib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage[]{todonotes}  % use [disable] to hide notes

@@

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  has a stabilizing effect down to about 4~\si{\kph}. We hypothesize that a
  stabilized bicycle will reduce the probability of falling. To test the
  motor's possible assistance during falls, we apply varying magnitude external
  handlebar perturbations to twenty-six participants who rode on a treadmill
  with the balance assist system activated and deactivated. The probability of
  recovering from a handlebar perturbation significantly increases when the
  balance assist is activated at a travel speed of 6~\si{\kph}. This positive
  effect is most prominent at and around the individual riders' perturbation
  resistance threshold. We conclude that use of a balance assist system in real
  world bicycling would reduce the number of falls that occur near riders'
  control authority limits.
}

@@

\section*{Keywords}
bicycle, fall, prevention, automatic control, stability

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
\citet{Whipple1899} not only derived the correct equations of motion of the
bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
can stabilize a bicycle. Much later, attempts at automatic roll stabilization
of single-track vehicles began in earnest after predictive motorcycle models
were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
influenced by Whipple's work and seems to be the first to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the early 1980s that \citet{Nagai1983}
successfully demonstrated balancing a robotic bicycle on a treadmill with both
steering control and a laterally moving mass. \citet{Ruijs1986} followed this
breakthrough by demonstrating an automatically balanced motorcycle and they did
so solely with a steering motor. Ruijs and Pacejka clearly showed that steer
torque driven by roll angle feedback stabilizes the capsize mode, by roll
angular rate feedback stabilizes the weave mode, and by steer angular rate
feedback stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle)
eigenmodes are defined in \citep{Sharp1971}.} They also showed how the control
gains must change with respect to vehicle speed for favorable control across
all speeds. This roll motion feedback enables the simplest controller that can
stabilize a single-track vehicle above a minimum speed when one is not
concerned with wobble instabilities. But Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, i.e no human rider was involved.

@@

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
free movement pre-fall. The participants practiced riding on the treadmill
until they indicated they were comfortable enough to have perturbations
applied. For most, this was less than a 10~\si{\minute} warm up. We then asked
the rider to ride for 90~\si{\second}, while attempting to maintain the
location of their front wheel on the center line of the treadmill as a baseline
measure before the perturbations. We then applied perturbations in random
directions (clockwise or counter-clockwise), starting at 20~\si{\newton} and
increasing the magnitude by 30~\si{\newton} until the participants fell. We
defined a ``fall'' on the treadmill by two criteria: 1) the rider removes their
foot from the pedal and places it on the ground or 2) the bicycle wheel exceeds
the width of the treadmill belt. Figure~\ref{fig:perturbation-sequence} shows
an example resulting motion from a perturbation. We logged the force magnitude
that caused the first fall to characterize that participant's
\emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

To illustrate the effect of the balance assist system on fall probability, we
will give an example of how the data collected during the experiments is used
to predict fall probability. We use Equation~\ref{eq:log-regress} and the
coefficients in Table~\ref{tab:freq-coefs-6}. For simplicity's sake, the
interaction effects are not included. For example, assuming that the mean
angular impulse $\bar{L}$ of all the perturbations applied to a participant is
100~\si{\newton} and the standard deviation $\sigma_{L}=15$. The centered and
scaled angular impulse can be calculated by subtracting $\bar{L}$ from the
applied angular impulse $L$, and dividing this by $\sigma_{L}$. The same
applies for the perturbation order $j$, initial roll angle $\phi_0$, and
initial steer angle $\delta_0$. If we take the coefficients estimated for
cycling at 6~\si{\kph}, the log-odds of falling can be calculated as follows:
\todo{Check this equation, where is \(\bar{L}\) 100N in the equation?}
%
\begin{equation}
  \begin{split}
    \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
    = & \quad
    \beta + \sum_{k=1}^{k=5}\alpha_k \frac{x_{ij}^{k}-\bar{x}_{ij}^{k}}{\sigma^{x^{k}}} \\
    = &
   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
    & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot s \\
    = &
    1.42 -0.64 \cdot s
  \end{split}
  \textrm{.}
\end{equation}
%
The state of the balance assist \(s\) is a binary variable. If the
balance assist is turned on, the log-odds that a fall occurs are decreased by
0.64. The odds and probability can be calculated:
%
\begin{align}
  \frac{p_{ij}}{1-p_{ij}} & = e^{1.42-0.64s} =
  e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53s \\
  p_{ij}|_{c=0} & = \frac{4.14}{1 + 4.14} = 0.81 \\
  p_{ij}|_{c=1} & = \frac{4.14\cdot0.53}{1 + 4.14\cdot{0.53}} = 0.69
\end{align}
%
Turning on the balance assist system reduces the probability that the
perturbation results in a fall from 81\% to 69\% in this case.

@@

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why the balance assist system did not have a
statistically significant effect at 10~\si{\kph}. As a bicycle travels at
higher speeds, the same perturbation magnitude causes larger lateral
deviations. At 10~\si{\kph} almost all falls were due to the bicycle exiting
the maximum width of the treadmill. If the same experiment was performed on an
infinitely wide plane, the riders may have recovered from more perturbations.
At 6~\si{\kph} the riders could often recover in the allotted treadmill width
due to the smaller lateral deviations. We believe our results are very much
dependent on the two modes of falling, i.e. exit the treadmill width or foot is
placed on the belt. On the other hand, Cycle paths are a similar width as the
treadmill, so rider's are often limited in width when recovering from a fall
thus exiting the treadmill width may be an appropriate measure for indicating a
fall.

>>>>

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage[natbib=true,style=numeric,sorting=none,giveninits=true]{biblatex}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\addbibresource{references.bib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage[]{todonotes}  % use [disable] to hide notes

@@

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  stabilizes the riderless bicycle down to about 3~\si{\kph}. We hypothesize
  that a stabilized bicycle will reduce the probability of falling. To test the
  motor's possible assistance during falls, we apply varying magnitude external
  handlebar perturbations to twenty-six participants who rode on a treadmill
  with the balance assist system both activated and deactivated. The
  probability of recovering from a handlebar perturbation significantly
  increases when the balance assist is activated at a travel speed of
  6~\si{\kph}. This positive effect is most prominent at and around the
  individual riders' perturbation resistance threshold. We conclude that use of
  a balance assist system in real world bicycling would reduce the number of
  falls that occur near riders' control authority limits.
}

@@

\section*{Keywords}
bicycle, fall prevention, automatic control, stability

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
\citet{Whipple1899} not only derived the correct equations of motion of the
bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
can stabilize a bicycle. Much later, attempts at automatic roll stabilization
of single-track vehicles began in earnest after predictive motorcycle models
were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
influenced by Whipple's work and seems to be the first to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the early 1980s that \citet{Nagai1983}
successfully demonstrated balancing a robotic bicycle on a treadmill with both
steering control and a laterally moving mass. \citet{Ruijs1986} followed this
breakthrough by demonstrating an automatically balanced motorcycle and they did
so solely with a steering motor. Ruijs and Pacejka showed that steer torque
driven by roll angle feedback stabilizes the capsize mode, by roll angular rate
feedback stabilizes the weave mode, and by steer angular rate feedback
stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes
are defined in \citep{Sharp1971}.} They also showed how the control gains must
change with respect to vehicle speed for favorable control across all speeds.
This roll motion feedback enables the simplest controller that can stabilize a
single-track vehicle above a minimum speed when one is not concerned with
wobble instabilities. But Ruijs and Pacejka's work was not particularly
concerned with low speed stability and their vehicle was fully automatic, i.e
no human rider was involved.

@@

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced riding on the
treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, while attempting to
maintain the location of their front wheel on the center line of the treadmill
as a baseline measure before the perturbations. We then applied perturbations
in random directions (clockwise or counter-clockwise), starting at
20~\si{\newton} and increasing the magnitude by 30~\si{\newton} until the
participants fell. We defined a ``fall'' on the treadmill by two criteria: 1)
the rider removes their foot from the pedal and places it on the ground or 2)
the bicycle wheel exceeds the width of the treadmill belt.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

To illustrate the effect of the balance assist system on fall probability, we
will give an example of how the data collected during the experiments is used
to predict fall probability. We use Equation~\ref{eq:log-regress} and the
coefficients in Table~\ref{tab:freq-coefs-6}. For simplicity's sake, the
interaction effects are not included. For example, assuming that the mean
angular impulse $\bar{L}$ of all the perturbations applied to a participant is
100~\si{\newton} and the standard deviation $\sigma_{L}=15$. The centered and
scaled angular impulse can be calculated by subtracting $\bar{L}$ from the
applied angular impulse $L$, and dividing this by $\sigma_{L}$. The same
applies for the perturbation order $j$, initial roll angle $\phi_0$, and
initial steer angle $\delta_0$. If we take the coefficients estimated for
cycling at 6~\si{\kph}, the log-odds of falling can be calculated as follows:
\todo{Check this equation, where is \(\bar{L}\) 100N in the equation?}
%
\begin{equation}
  \begin{split}
    \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
    = & \quad
    \beta + \sum_{k=1}^{k=5}\alpha_k \frac{x_{ij}^{k}-\bar{x}_{ij}^{k}}{\sigma^{x^{k}}} \\
    = &
   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 100\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
    & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot s \\
    = &
    1.59 -0.64 \cdot s
  \end{split}
  \textrm{.}
\end{equation}
%
The state of the balance assist \(s\) is a binary variable. If the
balance assist is turned on, the log-odds that a fall occurs are decreased by
0.64. The odds and probability can be calculated:
%
\begin{align}
  \frac{p_{ij}}{1-p_{ij}} & = e^{1.59-0.64c} =
  e^{1.59}\cdot e^{-0.64s} = 4.90 \cdot 0.53c \\
  p_{ij}|_{c=0} & = \frac{4.90}{1 + 4.90} = 0.83 \\
  p_{ij}|_{c=1} & = \frac{4.90\cdot0.53}{1 + 4.90\cdot{0.53}} = 0.72
\end{align}
%
Turning on the balance assist system reduces the probability that the
perturbation results in a fall from 83\% to 72\% in this case.

@@

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why the balance assist system did not have a
statistically significant effect at 10~\si{\kph}. As a bicycle travels at
higher speeds, the same perturbation magnitude causes larger lateral
deviations. At 10~\si{\kph} almost all falls were due to the bicycle exiting
the maximum width of the treadmill. If the same experiment was performed on an
infinitely wide plane, the riders may have recovered from more perturbations.
At 6~\si{\kph} the riders could often recover in the allotted treadmill width
due to the smaller lateral deviations. We believe our results are very much
dependent on the two modes of falling, i.e. exit the treadmill width or foot is
placed on the belt. On the other hand, cycle paths are a similar width as the
treadmill, so rider's are often limited in width when recovering from a fall
thus exiting the treadmill width may be an appropriate measure for indicating a
fall. While the treadmill simulates narrow cycle paths, real-world paths may
offer more lateral recovery space, providing riders with additional
opportunities to regain balance after a perturbation. However, testing in such
narrow conditions is still highly relevant, as system design and validation
should focus on extreme conditions like narrow paths, because even in wider
paths, obstacles such as parked cars or barriers can limit lateral space.
